A Large-Scale Study of Polymorphism in R

### 2019-04-06 paper/main.tex
stuff

In this section, we will present the results of our analysis at each of our type levels, describing the types of polymorphism that we have observed and shedding light on the frequency of polymorphic patterns in R.

Before we begin, we define the kinds of polymorphism we will consider.
In choosing how to express the polymorphicity of a function, there are two main options:
First, having several separate whole-function signatures, e.g. \D $\rightarrow$ \D and \C $\rightarrow$ \D, or alternatively expressing the polymorphism at the argument level with a union type, e.g. (\D $\cup$ \C) $\rightarrow$ \C.
We refer to the former as {\it per-function} or {\it function-level polymorphism}, and to the latter as {\it per-argument} or {\it argument-level polymorphism}.

We believe that both of these notions of polymorphism are valuable for different reasons.
Per-argument polymorphism highlights the type-based behavior of function arguments themselves, and helps to describe the type-based behavior of all R values.
For instance, we observe that many arguments have (\sC, \sF) as a signature, from which we deduce that many R programmers are taking advantage of functions such as \code{match.fun("fun\_name")} to call functions by name.
On the other hand, per-function polymorphism highlights how R programmers interact with functions as a whole, and allows us to discern more subtle patterns.
For example, observing two function signatures such as \D \to \D and \sF \to \sF may lead us to ascribe a more useful and precise type for the function than if we had simply collapsed the argument and return types (e.g., $X$ \to $X$ vs. \D $\cup$ \sF \to \D $\cup$ \sF).

The layout of this section is as follows:
In order, we present analysis results for {\bf L0}, {\bf L1}, and {\bf L2}, speaking to both per-argument and per-function polymorphism at each level.
Where appropriate, we digress to highlight interesting tidbits of tangential information which help us understand the style of code that R programmers write.
We begin with {\bf L0}.

@@

At {\bf L0}, the type system is mostly aligned with the types of the core language.

Across all of the arguments that we've observed, the vast majority are monomorphic.
Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.

@@

We see that doubles are the most common argument type, which is not entirely surprising as doubles are the backbone of any numeric data.
Next is lists, which seems reasonable given that lists are often used to store the aforementioned data.
Recall that in {\bf L0}, data frames will appear as lists.
The \ANY type is not an R type per se: we ascribe such a type to arguments which were {\it always} unevaluated across all calls.
As it happens, many R functions meta-program on some of their arguments, and this meta-programming does not force the argument promise, and if this meta-programming is always performed in every recorded call we will ascribe the \ANY type, as we cannot say what the argument {\it is} without possibly forcing a promise.
The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of meta-programming; we will discuss it in a little more detail shortly.
Interestingly, a polymorphic attribute signature \D, \l is among the top 10 overall---this is the vector and list polymorphism (shown in Figure~\ref{fig:listvec}) made manifest.

Now that we've seen the most common attribute-level signatures, we will turn our attention to function-level polymorphism.
Indeed, one of our goals is to generate useful function signatures, so we ask: {\it how many functions have more than one recorded signature?}
Figure~\ref{fig:L0funcounts} has the answer, with a breakdown by function.

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{L0_by_fun}
\caption{Record of Function Polymorphism in {\bf L0}. \LZEROPERCPOLY of functions are polymorphic.}
\label{fig:L0funcounts}\end{center}
\end{figure}

Here, we see that the vast majority of functions are monomorphic in L0.
That said, there are a few functions with a disproportionately large number of signatures.
One such function is the magrittr package's pipe \%$>$\%, which is used to implement a kind of currying (as per discussion in Section~\ref{sec:Impl})---here, \%$>$\% has nearly 100 unique signatures!
Many of the meta-programming functions in the Tidyverse are quite sophisticated, and are designed with the explicit goal of being very usable.
This usability maps to polymorphism in our analysis, so it's no surprise that a Tidyverse function features so prominently here.

Another function with many signatures is the data.table package's [ function, which is a reimplementation of the square bracket operator for data.tables.
This function has 16 formal arguments, and call configurations vary widely.
Essentially, [ is {\it the} main function for data.tables, and performs very different tasks based on which arguments are provided.
Two wildly different uses are:
\code{DT[i, j, by]}, where columns \code{i} and \code{j} are accessed and grouped by \code{by}, and \code{DT[x == "some" & y > 42]}, which selects all rows who's \code{x} column is the string \code{"some"} and who's \code{y} column is greater than \code{42}.
Given that, we find it unsurprising that the function is so polymorphic.

Having had a taste for function-level polymorphism, we will now explore common patterns in argument-level polymorphism.
For this, refer to Table~\ref{tab:L0toppoly}.

@@

Let's unpack some of the signatures in Table~\ref{tab:L0toppoly}:

	\item \D, \l (and \C, \l): L0 doesn't parameterize list types, and it's very likely that this signature describes vectors of doubles and lists of doubles.
	Coercing vectors to lists is trivial in R, and indeed many of the functions with such an argument signature do.
	The story is similar with vectors of characters and lists of characters.

@@

	\item \C, \sF: yet another puzzling signature, though its explanation is surprisingly straightforward:
	In R, programmers are free to look functions up {\it by name}, and indeed the frequently encountered {\tt UseMethod} function for dynamic class-based function dispatch is one such example.
	In short, this signature usually represents a function or a function name.

We have established and discussed some of the most common polymorphic attribute signatures, and have built up an understanding of argument-level polymorphism in the {\bf L0} type system.
This matches almost exactly with the language as it is today, as {\bf L0} is in (nearly) 1:1 correspondence with how types are currently represented in the language.
The next question we address is: {\it just how pervasive is this argument polymorphism?}
Refer to Table~\ref{tab:L0argcounts}.

\begin{table}[ht]\centering
\begin{tabular}{lrl}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 852K & 94\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\label{tab:L0argcounts}
\end{table}

@@

That said, L0 is not without limitations, and the main issues stem from a lack of adequate information, as the type system doesn't capture precise list types, and does not distinguish scalars from vectors.
The next type system we will explore, L1, tackles these issues by collecting more information to build more precise types.
While the L1 types are less ``in-tune'' with current R implementations, one can imagine designing an all-new R runtime which may benefit from the more specific information in L1, e.g. by not needing to vectorize scalar values, or designing a static analysis which uses precise list types to give programmers better information in an IDE.
L1 is discussed next.

%
%
%
%
\subsection{L1}

As we mentioned in Section~\ref{sec:L1def}, L0 is somewhat limited in its expressivity. 
To address this, we develop L1, which espouses a finer-grained approach to types. 
Major modifications from L0 include:

@@

From Table~\ref{tab:L1top10arg}, we see that all of the most frequent argument signatures are monomorphic even though we increased the precision of the types.
We see that scalars actually feature quite prominently, which is encouraging: this supports an idea that an explicit scalar annotation or type may be useful {\it for programmers}---indeed, such an annotation would also be helpful for compiler optimizations (e.g., scalar-only values need not have vectors allocated to store them).
We also see that matrices (\M{D}) are quite popular, a testament to the versatility of the data structure, as they can easily be converted to vectors and data frames.

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \D, \M{D} & 4848 & 0.5\% \\ 
  \l, \lT{D} & 1843 & 0.2\% \\ 
  \sC, \sD & 1590 & 0.2\% \\ 
  \M{D}, \M{I} & 1502 & 0.2\% \\ 
%  \l, \lT{list} & 1199 & 0.1\% \\ 
  \df, \M{D} & 1180 & 0.1\% \\ 
  \C, \D & 1019 & 0.1\% \\ 
  \sD, \sF & 1005 & 0.1\% \\ 
  \I, \sD & 997 & 0.1\% \\ 
  \sC, \sF & 991 & 0.1\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures in L1.}
\label{tab:toppolyL1}
\end{table}

@@

With an understanding of argument polymorphism in L1, we wonder: {\it how many argument positions are polymorphic in L1?}
Refer to Table~\ref{tab:argcountsL1}.

\begin{table}[ht]
\centering
\begin{tabular}{lrl}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 864K & 95.3\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L1.}
\label{tab:argcountsL1}
\end{table}

In spite of drastically increasing the precision of our types, we have actually {\it reduced} the amount of polymorphism.
This suggests that with a good notion of subtyping, precise type annotations won't cause an undue increase in the polymorphism of the language.

@@

%
%
\subsubsection{Scalars}
\label{sec:scalars}

>>>>

This section,  presents the results of our analysis at each of our type levels, describing the types of polymorphism that we have observed and shedding light on the frequency of polymorphic patterns in R.

Before we begin, we define the kinds of polymorphism we will consider.  In
choosing how to express the polymorphism of a function, we consider two
simple options: First, having the union of signatures, e.g. \D $\rightarrow$
\D and \C $\rightarrow$ \D, or a union types for each argument, e.g. (\D
$\cup$ \C) $\rightarrow$ \C.  We refer to the former as {\it function-level}
and the latter as {\it argument-level polymorphism}.

Per-argument polymorphism highlights the type-based behavior of function
arguments themselves, and helps to describe the type-based behavior of all R
values.  For instance, we observe that many arguments have (\sC, \sF) as a
signature, from which we deduce that many R programmers are taking advantage
of functions such as \code{match.fun("fun\_name")} to call functions by
name.  On the other hand, function-level polymorphism highlights how R
programmers interact with functions as a whole, and allows us to discern
more subtle patterns.  For example, observing two function signatures such
as \D \to \D and \sF \to \sF may lead us to ascribe a more useful and
precise type for the function than if we had simply collapsed the argument
and return types (e.g., $X$ \to $X$ vs. \D $\cup$ \sF \to \D $\cup$ \sF).

@@

At {\bf L0}, the type system is mostly aligned with the types of the core
language. Across all of the arguments that we have observed, the vast
majority are monomorphic.  Table~\ref{tab:L0top10arg} shows the breakdown of
the 10 most frequently occurring argument signatures.

@@

Doubles are the most common argument type, which is not entirely surprising
as doubles are the backbone of most statistical functions.  Next is lists,
which is reasonable as lists are used to build data frames and other widely
used data structures.  The \ANY type is not an R type per se: we ascribe
such a type to arguments which were {\it always} unevaluated across all
calls.  As it happens, many R functions meta-program on some of their
arguments, and this meta-programming does not force the argument promise,
and if this meta-programming is always performed in every recorded call we
will ascribe the \ANY type, as we cannot say what the argument {\it is}
without possibly forcing a promise.  The magrittr package's pipe \%$>$\%
function is an example of a function which does meta-programming; we will
discuss it shortly.  Interestingly, a polymorphic attribute signature \D, \l
is among the top 10 overall---this is the vector and list polymorphism made
manifest.

We turn our attention to function-level polymorphism.  Indeed, one of our
goals is to generate useful function signatures, so we ask: {\it how many
  functions have more than one recorded signature?}
Figure~\ref{fig:L0funcounts} has the answer, with a breakdown by function.
Most functions are monomorphic.  That said, a few functions have
disproportionately large number of signatures.  One such function is the
magrittr package's pipe \%$>$\%, which is used to implement a kind of
currying. It has nearly 100 unique signatures!

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{L0_by_fun}
\caption{Record of Function Polymorphism in {\bf L0}; \LZEROPERCPOLY of
  functions are polymorphic}
\label{fig:L0funcounts}\end{center}
\end{figure}

Another function with many signatures is the data.table package's \code{[}
  function, which is a reimplementation of the square bracket operator for
  selecting values in a data.table.  This function has 16 formal arguments,
  and call configurations vary widely.  Essentially, the function is {\it
    the} main function for data.tables, and performs very different tasks
  based on which arguments are provided.  Two wildly different uses are:
  \code{dtT[i,j,by]}, where columns \code{i} and \code{j} are accessed and
  grouped by \code{by}, and \code{dt[x=="some" & y>42]}, which selects
  all rows who's \code{x} column is the string \code{"some"} and who's
  \code{y} column is greater than \code{42}.  Given that, we find it
  unsurprising that the function is so polymorphic.

@@

Having had a taste for function-level polymorphism, we will now explore
common patterns in argument-level polymorphism.  For this, refer to
Table~\ref{tab:L0toppoly}.
Let's unpack some of the signatures in Table~\ref{tab:L0toppoly}:

\begin{itemize}
\item \D, \l (and \C, \l): L0 doesn't parameterize list types, and it's very likely that this signature describes vectors of doubles and lists of doubles.
	Coercing vectors to lists is trivial in R, and indeed many of the functions with such an argument signature do.
	The story is similar with vectors of characters and lists of characters.

@@

	\item \C, \sF: yet another puzzling signature, though its explanation is surprisingly straightforward:
	In R, programmers are free to look functions up {\it by name}, and indeed the frequently encountered {\tt UseMethod} function for dynamic class-based function dispatch is one such example.
	In short, this signature usually represents a function or a function name.
\end{itemize}

We have established and discussed some of the most common polymorphic
attribute signatures, and have built up an understanding of argument-level
polymorphism in the {\bf L0} type system.  This matches almost exactly with
the language as it is today, as {\bf L0} is in (nearly) 1:1 correspondence
with how types are currently represented in the language.  The next question
we address is: {\it just how pervasive is this argument polymorphism?}
We find that 94\% of arguments are monomorphic.

@@

That said, L0 is not without limitations, and the main issues stem from a
lack of adequate information, as the type system doesn't capture precise
list types, and does not distinguish scalars from vectors.

%
%
\subsection{L1}

L1 espouses a finer-grained approach to types.  Major modifications from L0
include:

@@

From Table~\ref{tab:L1top10arg}, we see that all of the most frequent
argument signatures are monomorphic even though we increased the precision
of the types.  We see that scalars actually figure quite prominently, which
is encouraging: this supports an idea that an explicit scalar annotation or
type may be useful {\it for programmers}---indeed, such an annotation would
be helpful for compiler optimizations (e.g., scalar-only values need not
have vectors allocated to store them).  We also see that matrices (\M{D})
are quite popular, a testament to the versatility of the data structure, as
they can easily be converted to vectors and data frames.

@@

\begin{table}[ht]\centering\begin{tabular}{lrl}  \hline
Signature & Count & \% of All Signatures \\   \hline
  \D, \M{D} & 4848 & 0.5\% \\ 
  \l, \lT{D} & 1843 & 0.2\% \\ 
  \sC, \sD & 1590 & 0.2\% \\ 
  \M{D}, \M{I} & 1502 & 0.2\% \\ 
%  \l, \lT{list} & 1199 & 0.1\% \\ 
  \df, \M{D} & 1180 & 0.1\% \\ 
  \C, \D & 1019 & 0.1\% \\ 
  \sD, \sF & 1005 & 0.1\% \\ 
  \I, \sD & 997 & 0.1\% \\ 
  \sC, \sF & 991 & 0.1\% \\    \hline
\end{tabular}
\caption{Top polymorphic argument signatures in L1.}\label{tab:toppolyL1}
\end{table}

@@

With an understanding of argument polymorphism in L1, we wonder: {\it how
  many argument positions are polymorphic in L1?}  95.3\% of arguments are
monomorphic.  In spite of drastically increasing the precision of our types,
we have actually {\it reduced} the amount of polymorphism.  This suggests
that with a good notion of subtyping, precise type annotations won't cause
an undue increase in the polymorphism of the language.

@@

%
%
\subsubsection{Scalars}\label{sec:scalars}
